Pop-up cycleways and pedestrian spaces: a peak into post Covid-19 transport futures

### 2020-05-09 conversation.Rmd
Conversation piece

The Covid-19 pandemic has transformed all sectors of the economy, not least transport. Long distance travel has been dramatically altered, as [demand plummets](https://osf.io/preprints/socarxiv/v3g5d/) and [airline companies go bust](https://arxiv.org/abs/2004.08460). Yet short distance travel patterns have also changed greatly. Forced to stay close to home, many people have started walking and cycling more. From a health and well-being perspective, this may be a more important shift, especially in the long term as coronavirus measures are gradually eased.

Local and national governments worldwide are still working out how best to respond to these changes.
Hundreds of kilometres of temporary infrastructure being put in place in dozens of cities worldwide, from [Park Lane in London](https://www.forbes.com/sites/carltonreid/2020/05/06/pop-up-cycleway-for-park-lane-as-tenfold-increase-in-cycling-predicted-for-london/#48aad0705b1e) to [Martin Luther King Jr Drive in Philadelphia](https://www.theguardian.com/world/2020/apr/11/world-cities-turn-their-streets-over-to-walkers-and-cyclists ""), [Auckland](https://www.roadtraffic-technology.com/news/nz-govt-extra-wide-footpaths/) to [Mexico City](https://bicitekas.org/sites/default/files/public/manuales/propuesta_ciclovias_temporales_cdmxvfinal.pdf), never has there been such a need for affordable, healthy and socially isolated transport that does not create the fumes that emerging evidence links to coronavirus susceptibility. With the rush to act, there is a great need for evidence of *where* new interventions should be prioritised. As with the medical science, research is needed now. Methods developed to identify locations of high walking and cycling potential can help ensure that the ‘pop-up’ infrastructure that goes in now is effective, safe, and placed where it is most needed.

>>>>

Around the world, as nations prepare to emerge from months of Covid-19 induced lockdown, life cannot simply return to previous norms. The pandemic has transformed all sectors of the economy, not least transport. Long distance travel has been dramatically altered, as [demand plummets](https://osf.io/preprints/socarxiv/v3g5d/) and [airline companies go bust](https://arxiv.org/abs/2004.08460). Yet short distance travel patterns have also changed greatly. Forced to stay close to home, many people have started walking and cycling more. From a health and well-being perspective, this may be a more important shift, especially in the long term as lockdown measures are gradually eased.

In the coming weeks and months, active travel will play a vital role in enabling the reopening of many countries. As travel increases, we must ensure that safe, socially distanced walking and cycling remains possible. Maintaining social distancing on public transport could severely reduce the capacity of rail and bus networks, and there is a risk that this may lead to a rapid rise in road traffic. Without careful management, increased congestion could deter others from the streets, especially as [ record bicycle sales ] suggest that many people have taken advantage of lockdown conditions to take up or return to cycling. These people may have less experience in dealing with high volumes of road traffic.

Active travel has extensive [ social and environmental benefits ]. Physical activity is associated with XX health conditions….. mental health…..isolation...

Evidence is beginning to emerge of [ potential links ] between air pollution and Covid-19 lethality, not least because many health conditions such as asthma and xxx are [ worsened by poor air quality ]. Indeed, it is estimated that global improvements in air quality in recent months have led to [ xx thousand fewer deaths ] from conditions such as these. This makes it all the more important to restrict any rise in road traffic and to encourage more beneficial means of travel.

Local and national governments worldwide are still working out how best to respond to these challenges. On Saturday 9th May, British Transport Secretary Grant Shapps announced a £2 billion investment in measures to promote walking and cycling. This follows a £10 million commitment from the Scottish Government and action from Wales. Hundreds of kilometres of temporary infrastructure is being put in place in dozens of cities worldwide, from [Milan] to [Martin Luther King Jr Drive in Philadelphia](https://www.theguardian.com/world/2020/apr/11/world-cities-turn-their-streets-over-to-walkers-and-cyclists ""), [Auckland](https://www.roadtraffic-technology.com/news/nz-govt-extra-wide-footpaths/) to [Mexico City](https://bicitekas.org/sites/default/files/public/manuales/propuesta_ciclovias_temporales_cdmxvfinal.pdf). Never has there been such a need for affordable, healthy and socially distanced transport that does not create the fumes that [emerging evidence] links to coronavirus susceptibility.

In terms of practical steps to encourage walking and cycling, a range of simple interventions can be undertaken, initially on a temporary basis. Roads can be closed to motor vehicles, or physical barriers can be used to enable [ filtered permeability ], in which access for residents is retained but modal filters enable routes for walking and cycling that are faster and more direct than the routes for motor vehicles. This is well established in [ the London Borough of Hackney ] where attractive barriers such as planters have been used to good effect. Another option involves keeping the roads open while reallocating excess roadspace to widen pavements or introduce pop-up cycleways.

These measures can all be undertaken on a temporary basis, but if action is prioritised in places where there is a *long-term need* for cycle improvements, there is a greater chance that people will demand they become permanent. This is what happened with the experimental cycleway in Torrington Place in London, where a 2 way road for motor vehicles has been converted into a thriving 2 way cycleway, enabling social distancing and zero emissions transport for keyworkers in central London. [Paris plans].

With the rush to act, there is a great need for evidence of *where* new interventions should be prioritised. As with the medical science, research is needed now. Methods developed to identify locations of high walking and cycling potential can help ensure that the ‘pop-up’ infrastructure that goes in is effective, safe, and placed where it is most needed. Agile research can respond to rapidly changing conditions, as government priorities shift and different sectors of the economy reopen.

Building on the previous projects such as the [ Propensity to Cycle Tool ] and the [ Cycling Infrastructure Prioritisation Toolkit ], researchers from the University of Leeds have investigated where roadspace could be reallocated for pop-up cycleways. They have mapped out the top locations for intervention in ten major cities across England, with the intention of rolling out these methods across the country in the near future.

The methods have focused on road capacity and cycling potential, identifying roads where a ‘spare lane’ could be repurposed for cycling. Cycle potential is assessed based on travel to work data and estimates of cycling uptake under the UK Government’s national target of doubling cycling levels by 2025, with the length of road segments also being used to filter the top schemes.

The use of open data and publicly accessible models is a key aspect of the research. Given the global nature of the challenges we face, open and collaborative research is vital so that methods can be applied by as wide a user base as possible. The potential for international application can be seen for example in [ WHO report ] modelling cycling uptake in cities in low income countries. Within the UK, [ 50+ Local Authorities ] are now using the PCT to inform their strategic cycle network plans.

Maps (see below) show the recommendations for where to introduce pop-up cycleways. With the situation rapidly developing, some of the roads shown here, such as [ Princess Street in Manchester ], have already been identified by Local Authorities as targets for intervention.
